---
title: "Practical Machine Learning Course Project"
author: "Ong Kwee Hian"
date: "Saturday, August 22, 2015"
output: html_document
---


## Introduction
In this project, the goal is to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants who were asked to perform barbell lifts correctly and incorrectly in 5 different ways as trainng data set to develop a model, and finally use it to predict the manner in which they did the exercise in the 20 test data set observations.

## Load and prepare/clean Training data 

Attempt to first investigate the characteristics of raw training data set.
```{r warning=FALSE}
library(lattice)
library(ggplot2)
library(caret)

rawTrainData <- read.csv("pml-training.csv")  # Training data set
rawTestData <- read.csv("pml-testing.csv")    # Test data set

str(rawTrainData)
```
There a quite a large number of NA and also variable not useful for analysis. To clean away unnecessary variables, i will remove variables with nearly zero variance, variables that has NA, and variables that are irrelevant for prediction e.g. user_name, time stanp, etc. 
```{r}
# remove near zero variance variables
nzv <- nearZeroVar(rawTrainData)
rawTrainData <- rawTrainData[, -nzv]

# remove NA variables
NAs <- apply(rawTrainData, 2, function(x) { sum(is.na(x)) })
rawTrainData <- rawTrainData[, which(NAs == 0)]

# remove first five irrlevant variables (X, user_name, raw_timestamp_part_1, raw_timestamp_part_2, cvtd_timestamp)
rawTrainData <- rawTrainData[, -(1:5)]

str(rawTrainData)
```

To be able to estimate the out-of-sample error, I will make use of this training set by splitting the Full training data set (rawTrainData) into 2 smaller sets for use, one for training purpose (rawTrainData1) and the other for validation purpose (rawTrainData2):

```{r}
set.seed(082015)
inTrain <- createDataPartition(y=rawTrainData$classe, p=0.7, list=F)
rawTrainData1 <- rawTrainData[inTrain, ]    # For training
rawTrainData2 <- rawTrainData[-inTrain, ]   # For testing
```

## Build Model 

Applying Random Forest model on rawTrainData1, and "train" using 3-fold cross-validation to select optimal tuning parameters for acceptable performance for the model.

```{r}
# Use 3-fold cross-validation to select optimal tuning parameters
modControl <- trainControl(method="cv", number=3, verboseIter=FALSE)

# fit model on rawTrainData1
modFit <- train(classe ~ ., data=rawTrainData1, method="rf", trControl=modControl)

# print tuning parameters choosen in final model
modFit$finalModel
```

The result return is 500 trees and try 27 variables at each split.

## Model Evaluation and Selection

This model is used to predict the label ("classe") in rawTrainData2, and confusion matrix is used to compare the predicted versus the actual labels:

```{r}
# Predict classe label in validation set (rawTrainData2) using model
predictions <- predict(modFit, newdata=rawTrainData2)

# Print confusion matrix to understand out-of-sample error
confusionMatrix(rawTrainData2$classe, predictions)
```

The accuracy is 99.8%, thus my predicted accuracy for the out-of-sample error with cross validation is 0.2%, which is very good.


## Re-training the Model using Full training data

As the result is very good, Random Forests will be used to predict labels for the test set (20 test cases) provided. I will now retrain the model using the Full training set (rawTrainData) before predicting on the test set given. 

```{r}
# re-model using full training set (rawTrainData)
modControl <- trainControl(method="cv", number=3, verboseIter=F)
modFit <- train(classe ~ ., data=rawTrainData, method="rf", trControl=modControl)

# print tuning parameters choosen in final model
modFit$finalModel
```

## Using the model for Test Set Predictions

The new model fit is applied on activity monitors in rawTestData to predict the activity quality label, and output to files:

```{r eval=FALSE}
# predict on test set
predictions <- predict(modFit, newdata=rawTestData)

# convert predictions to list of vector
predictions <- as.character(predictions)

# create function to write predictions to files
pml_write_files <- function(x) {
    n <- length(x)
    for(i in 1:n) {
        filename <- paste0("problem_id_", i, ".txt")
        write.table(x[i], file=filename, quote=F, row.names=F, col.names=F)
    }
}

# create prediction files to submit
pml_write_files(predictions)
```

The created  result file input are used for project submission.
As the model take a long time to generate, the model output are only show in html file.

